### 2024-04-05 doc/15yrsChange.Rmd
integrated edits to results from google doc

Pests and pathogens; temperate forests; carbon sequestration; white-tailed deer; ecosystem processes;

@@

All sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals. However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhitibed significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

@@

For the first time since the establishment of  the SCBI ForestGEO plot, biomass losses to mortality outpaced biomass gains from tree growth (Figure 3a). Living aboveground C stocks (AGB) decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a). In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

@@

 Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c). Plot-wide AWG decreased by `r (awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013 * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval. Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^). Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d). AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.

Deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5). The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^. In particular, the recruitment rate of *Carya* spp. and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density. The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory. However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp. was lower. Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa. Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability. *Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).

@@

There were significant areas of decreasing AGB and suppressed regeneration of canopy species. Specifically, 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory. Only 21.7% of the plot exhibits an understory with high AGB of species capable of reaching the canopy. Within the region characterized by high deer density and high canopy vulnerability 84% of the area has lost canopy biomass, only 5.4% has an understory containing high AGB of canopy taxa, and 0% of area exhibits both. In contrast, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB. Furthermore, 70.7% of this same region has high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration.

![**_Figure `r regen_failure`. Characterizing the SCBI ForestGEO plot by changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023. (b) The % area of each region that falls within each classification.](display/Figure6.jpeg)

>>>>

Pests and pathogens; temperate forests; carbon sequestration; white-tailed deer; ecosystem processes; invasive species

@@

Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3b). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals. However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

@@

Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure 3a). For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure 2), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a). In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

@@

 Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c). Plot-wide AWG decreased by `r round((awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013,1) * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval. Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^). Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d). AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.

Consistent with hypotehsis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5). The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^. In particular, the recruitment rate of *Carya* spp. and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density. The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory. However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp. was lower. Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa. Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability. *Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).

@@

There were significant areas of suppressed regeneration of canopy species, often coincident with decreasing AGB. Only 21.7% of the plot (mostly inside the deer exlosure) exhibited an understory with high AGB of species capable of reaching the canopy. Within the plot areas characterized by high deer density and high canopy vulnerability, 84% of the area had lost canopy biomass, only 5.4% had an understory containing high AGB of canopy taxa, and 0% of area exhibited both a healthy understory and increasing AGB. In contrast, 70.7% of the plot area with low deer density and low canopy vulnerability had high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration. Furthermore, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB.

![**_Figure `r regen_failure`. Map of changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023 within the SCBI ForestGEO plot._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023. (b) The % area of each region that falls within each classification.](display/Figure6.jpeg)
